Move Schedules: Fast persistence computations in sparse dynamic settings \thanks{Submitted to the editors DATE. \funding{This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758.} }

### 2021-08-21 move_schedules.tex
Updated for SIAM format

\documentclass[10pt]{article}
\usepackage[margin=0.75in]{geometry}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{mathtools}
\usepackage{url}
\usepackage{enumitem}
\usepackage{setspace}
\usepackage{scalefnt}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{algorithm}
%\usepackage{algorithm2e}
%\usepackage{algorithmicx}
\usepackage[noend]{algpseudocode}
\usepackage{bbm}
\usepackage{graphicx}

@@

@@

\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
% \newcommand{\CommentX}[1]{\unskip~\%~#1~\%}

@@

%Unraveling
\title{\vspace{-2.0em} Move Schedules: \\ Fast persistence computations in sparse dynamic settings\thanks{This work was partially supported by the National Science Foundation through grants CCF-2006661
and CAREER award  DMS-1943758.}\vspace{-0.5em}}
\author{Matt Piekenbrock\thanks{Department of Computational Mathematics, Science \& Engineering, Michigan State University}\;\; and Jose A. Perea$^\dagger$\thanks{Department of Mathematics, Michigan State University}}
\date{}

\begin{document} \vspace{-4em} \maketitle
\begin{abstract}
	The standard procedure for computing the persistent homology of a filtered simplicial complex is the matrix reduction algorithm. Its output is a particular decomposition of the total boundary matrix, from which the persistence diagrams and generating cycles can be derived. 
	Persistence diagrams are known to vary continuously with respect to their input; this motivates the algorithmic study of persistence computations for time-varying filtered complexes. Computationally, simulating persistence dynamically can be reduced to maintaining a valid decomposition under adjacent transpositions in the filtration order. In practice, the quadratic scaling in the number of  transpositions often makes this maintenance procedure slower than simply computing the decomposition from scratch, effectively limiting the application of dynamic persistence to relatively small data sets. In this work, we propose a coarser strategy for maintaining the decomposition over a discrete 1-parameter family of filtrations. Our first result is an analysis of a simple linear-time strategy which reduces the number of column operations needed to simulate persistence across a fixed homotopy by at most a factor of 2. We then show a modification of this technique which maintains only a sublinear number of valid states, as opposed to a quadratic number of states, and we provide tight lower bounds for this technique.
	Finally, we provide empirical results suggesting that the decrease in operations needed to compute diagrams across a family of filtrations is proportional to the difference between the expected quadratic number of states, and the proposed sublinear coarsening.
\end{abstract}

@@

 \subsection{A Motivating Example and Contributions}\label{sec:motivation} 
As a motivating example, we describe a simple experiment illustrating why the vineyards algorithm does not always yield an efficient method in many time-varying applications. 
Consider a series of grayscale images (i.e. a video) depicting a fixed-width annulus expanding about the center of a $9 \times 9$ grid, where scale of the annulus is parameterized by a single parameter $\tau$. Thus, the data set is a 1-parameter family of images. Each image contains the same number of pixels whose intensities---these are determined by the parameter---vary with time. For each image, we build a filtered simplicial complex using the \emph{Freudenthal} triangulation, with the ordering of inclusions obtained by lower stars.  
 \begin{figure*}[t]
 	\centering
 	\includegraphics[width=0.36\textwidth, height = 1.65in]{circle_vineyards.png}
 	\includegraphics[width=0.63\textwidth]{circle_complex.png}
 	\caption{Grayscale image data set depicting an expanding annulus (top right), each image's corresponding lower star filtration (bottom right), and the cumulative number of column operations required to compute persistence for this time-varying filtration (left). The persistent Betti numbers for each filtration from left to right are: $\beta_1 = (1,0,0), \beta_2 = (1,1,0), \beta_3 = (1,1,0), \beta_4 = (1,1,0), \beta_5 = (4,0,0)$. Observe computing $10$ diagrams independently at evenly spaced time points (the green line) both captures these major topological changes and is the most computationally efficient approach shown.}
 	\label{fig:vineyards}
 \end{figure*}
Snapshots of the image data and its corresponding time-varying filtration is depicted in Figure~\ref{fig:vineyards} (top right and bottom right, respectively), which we use as a baseline for comparing the standard algorithm $\texttt{pHcol}$ with vineyards. Note the size of the filtration is fixed--no simplices are ever added or removed over time. 
There are two events which dramatically alter the persistence diagrams extracted from these filtrations: the first occurs when the central connected component splits to form a cycle, and the second occurs when the annulus approaches the outer dimensions of the grid, splitting into four components. As a consequence, in this example, only a few persistence diagrams are needed to capture the major topological changes over time.

@@

\textbf{Our results and approach are as follows:} First, we leverage work of Busaryev et. al.~\cite{busaryev2010tracking} to include a coarser type of operation to the dynamic setting, called \emph{moves}. We give a tight lower bound on the number of moves needed to perform an arbitrary permutation to the $R = D V$ decomposition, and give a proof of optimality by a reduction to the permutation edit distance problem.
We also give worst-case bounds in expectation as well as efficient algorithms for achieving these bounds---both of which are derived from a reduction to the Longest Increasing Subsequence (LIS) problem. 
This reduction yields an efficient algorithm for generating sequences of moves $( s_1, s_2, \dots, s_d )$, which we call \emph{schedules}. 
We investigate the feasibility of optimizing the cost 
(i.e., number of column operations) of these schedules directly, including the existence of a greedy-type approach, which we show can lead to arbitrarily bad behavior. 
In light of these results, we give an alternative proxy-objective to minimize, provide bounds justifying its relevance to the original objective, and give an efficient $O(d \log d)$ algorithm for approximately solving this proxy minimization. 
A performance comparison with other reduction-based persistent homology computations is given, wherein our approach is demonstrated to be an order of magnitude more efficient than existing approaches in a few common sparse dynamic settings, such as flock simulation and 2D persistence computation. We conclude with a discussion of situations where our strategy is most applicable and future directions for research.

@@

 The main output of the reduction algorithm is a matrix decomposition $R = D V$, where the persistence diagram is encoded in $R$. 
 Given a  filtration $K$ with $m$ simplices and maximal dimension $d$, 
 one assembles the elementary boundary chains $\partial(\sigma)$ as columns ordered according to the filtration order, forming the \emph{filtration boundary matrix} $D$. 
 The reduction algorithm can  compute the persistent homology for all dimensions up to $d - 1$.
 In this case,  $D$ is a square matrix of size $m \times m$. Alternatively, a single dimension $i \leq d - 1$ can be computed, in which case one reduces a pair of matrices $D_i, D_{i+1}$, where the former has dimension $d_{i-1} \times d_i$ and the latter $d_{i} \times d_{i+1}$. If one is interested in a particular dimension of homology then the latter approach is preferred. However, the algorithmic description of the algorithms is simpler when considering the full matrix, so $D$ will always be $(m \times m)$ in the descriptions that follow. All algorithms discussed generalize to both situations. We give an example using the matrix pair $(R_1, V_1)$  below. 
\textbf{Example 2.1: Reduction} Consider a triangle with vertices $u,v,w$ and edges $a = (u,w)$, $b = (v,w)$, $c = (u, v)$ whose filtration order, read as inclusions from left to right, is given as $K = (u, v, w, a, b, c)$. The reduction algorithm begins by setting $R = D$ and $V = I$, followed by left-to-right column operations until invariant (2) below is satisfied.
If the $i$-th column of $R$ is nonzero, then $\mathrm{low}_R(i)$ 
is the row index of its lowest nonzero entry.
Using $\mathbb{Z}_2$ coefficients and homology in dimension 1 to simplify the presentation, the steps look at follows:
	\makeatletter\setlength\BA@colsep{2.8pt}\makeatother
  		u \kern-0.1em  & \; 1 &    &  1 \; \topstrut \\
  		v \kern-0.1em & \;     &  1 & 1 \; \\
  		w \kern-0.1em & \; 1 &  1 &    \; \botstrut \\
	\; , \;
        a \kern-0.1em  & \; 1 &  &  \; \topstrut \\
  		b \kern-0.1em & \;  & 1 & \; \\
  		c \kern-0.1em & \;  &  & 1 \; \botstrut \\
	\quad \xrightarrow{} \,
  		u \kern-0.1em  & \; 1 & 1 & 1 \; \topstrut \\
  		v \kern-0.1em & \;     & 1 & 1 \; \\
  		w \kern-0.1em & \; 1 &    &    \; \botstrut \\
	\; , \;
  		a \kern-0.1em  & \; 1 & 1 &    \; \topstrut \\
  		b \kern-0.1em & \;     & 1 &    \; \\
  		c \kern-0.1em & \;     &    & 1 \; \botstrut \\
	\quad \xrightarrow{} \,
  		u \kern-0.1em  & \;     & 1  &  \; \topstrut \\
  		v \kern-0.1em &  \; 1  & 1  &  \; \\
  		w \kern-0.1em & \; 1  &     &  \; \botstrut \\
	\; , \;
  		b \kern-0.1em  & \; 1 & 1 & 1 \; \topstrut \\
  		a \kern-0.1em & \;     & 1 & 1 \; \\
  		c \kern-0.1em & \;     &    & 1 \; \botstrut \\
	\vspace*{-1em}

@@

\begin{proposition}[Persistent Pairing~\cite{zomorodian2005computing}]
	Let $R = D V$, where $R$ is in reduced form and $V$ is full-rank upper triangular (invariants (1) and (2)  hold). Then the pairing $(\sigma_i, \sigma_j)$ defined by $i = \mathrm{low}_R(j)$ is a \emph{creator/destroyer (persistent) pair}. While the matrices $R,V$ satisfying (1) and (2) are not unique in general,  the persistent pairs $(i,j)$ are unique and define the persistence diagrams of $K$.
\end{proposition}

@@

@@

	\makeatletter\setlength\BA@colsep{2.8pt}\makeatother

@@

  		u \kern-0.1em  & \; 1 & 1 &  \; \topstrut \\
  		v \kern-0.1em & \;  & 1 &  \; \\
  		w \kern-0.1em & \; 1 &  &  \; \botstrut \\

@@

	\, , 

@@

        a \kern-0.1em  & \; 1 & \tikz[baseline]{
            \node[fill=orange!20,anchor=base] (t1)
            {$ 1 $};
        } & 1 \; \topstrut \\
  		b \kern-0.1em & \;  & 1 & 1 \; \\
  		c \kern-0.1em & \;  &  & 1 \; \botstrut \\

@@

	\hspace{-0.5em}

@@

  		u \kern-0.1em  & \; 1 &  &  \; \topstrut \\
  		v \kern-0.1em & \;  & 1 &  \; \\
  		w \kern-0.1em & \; 1 & \tikz[baseline]{
            \node[fill=red!20,anchor=base] (t1)
            {$ 1 $};
        } &  \; \botstrut \\

@@

	\, , 

@@

  		a \kern-0.1em  & \; 1 &  & 1 \; \topstrut \\
  		b \kern-0.1em & \;  & 1 & 1 \; \\
  		c \kern-0.1em & \;  &  & 1 \; \botstrut \\

@@

	\hspace{-0.5em}

@@

  		u \kern-0.1em  & \;  & 1 &  \; \topstrut \\
  		v \kern-0.1em & \; 1 &  &  \; \\
  		w \kern-0.1em & \; 1 & \tikz[baseline]{
            \node[fill=red!20,anchor=base] (t1)
            {$ 1 $};
        } &  \; \botstrut \\

@@

  		b \kern-0.1em  & \; 1 &  & 1 \; \topstrut \\
  		a \kern-0.1em & \;  & 1 & 1 \; \\
  		c \kern-0.1em & \;  &  & 1 \; \botstrut \\
	\hspace{-0.5em}
  		u \kern-0.1em  & \;  & 1 &  \; \topstrut \\
  		v \kern-0.1em & \; 1 & 1 &  \; \\
  		w \kern-0.1em & \; 1 &  &  \; \botstrut \\
	\, ,
  		b \kern-0.1em  & \; 1 & 1 & 1 \; \topstrut \\
  		a \kern-0.1em & \;  & 1 & 1 \; \\
  		c \kern-0.1em & \;  &  & 1 \; \botstrut \\

@@

If one has a pair of filtrations $(K_0, K_1)$ each with $m$ simplices, and we assume that each simplex $\sigma_i$ switches its relative ordering with another simplex $\sigma_j$ at most once in some ordered schedule of transpositions, then the number of intermediate persistence diagrams is bounded above by $O(m^2)$ using transpositions, whereas with move operations this bound clearly is at most $O(m)$.  
\noindent \textbf{Example:} We continue the example used in sections~\ref{sec:vineyards} and~\ref{sec:reduction} to illustrate moves. Consider moving the edge $a$ to the position taken by the edge $c$ in the filtration. The donor columns introduced in~\cite{busaryev2010tracking} are shown on the left side of each matrix. Note that using vineyards, the equivalent permutation to the decomposition would require $4$ column operations on both $R_1$ and $V_1$, respectively, whereas a single move operation accomplishes this permutation using only $2$ column operations per matrix. 
	\makeatletter\setlength\BA@colsep{2.8pt}\makeatother
	\begin{blockarray}{cc}{d_R}
	\kern-0.2em & a \\
		\begin{block}{c[c]}
  		u \kern-0.1em  & \; 1 \; \topstrut \\
  		v \kern-0.1em & \;  \; \\
  		w \kern-0.1em & \; 1 \; \botstrut \\
	\; \;
  		u \kern-0.1em  & \; 1 & 1 &  \; \topstrut \\
  		v \kern-0.1em & \;  & 1 &  \; \\
  		w \kern-0.1em & \; 1 &  &  \; \botstrut \\
\; \to
\begin{blockarray}{cc}{d_R}
	\kern-0.2em & b \\
		\begin{block}{c[c]}
  		u \kern-0.1em  & \; 1 \; \topstrut \\
  		v \kern-0.1em & \;  1 \; \\
  		w \kern-0.1em & \; \; \botstrut \\
	\; \;
  		u \kern-0.1em  & \; 1 &    &  \; \topstrut \\
  		v \kern-0.1em  & \;    & 1 &  \; \\
  		w \kern-0.1em & \; 1 & 1 &  \; \botstrut \\
\; \to
	\begin{blockarray}{cc}{d_R}
	\kern-0.2em & c \\
		\begin{block}{c[c]}
  		u \kern-0.1em  & \; \hphantom{1} \; \topstrut \\
  		v \kern-0.1em & \;  \hphantom{1} \; \\
  		w \kern-0.1em & \; \hphantom{1} \; \botstrut \\
	\; \; 
  		u \kern-0.1em  & \; 1   &    &   1  \; \topstrut \\
  		v \kern-0.1em & \;       &  1  &  1  \; \\
  		w \kern-0.1em & \; 1   &   1   &     \; \botstrut \\
\; \xrightarrow{P}
\begin{blockarray}{cc}{d_R}
	\kern-0.2em & c \\
		\begin{block}{c[c]}
  		u \kern-0.1em  & \; \hphantom{1} \; \topstrut \\
  		v \kern-0.1em & \; \hphantom{1} \; \\
  		w \kern-0.1em & \; \hphantom{1} \; \botstrut \\
\; \;
	\kern-0.2em & b & c & a  \\
  		u \kern-0.1em  & \;  & 1 &  1\; \topstrut \\
  		v \kern-0.1em & \; 1 & 1 &  \; \\
  		w \kern-0.1em & \; 1 &  & 1 \; \botstrut \\
\; \xrightarrow{d_R}
	\kern-0.2em & b & c & a  \\
  		u \kern-0.1em  & \;  & 1 &  \; \topstrut \\
  		v \kern-0.1em & \; 1 & 1 &  \; \\
  		w \kern-0.1em & \; 1 &  &  \; \botstrut \\

@@

\begin{figure}[t]
	\centering
	\includegraphics[width=0.95\textwidth]{crocker_combo_1.png}
	\caption{An example of a crocker plot (right) depicting the evolution of dimension $p = 1$ Betti curves over time. The green $\mathrm{X}$ marks correspond chronologically to the complexes (left), in row-wise order. The large orange and purple areas depict $1$-cycles persisting in space (the y-axis) and time (x-axis). }\label{fig:crocker1}
\end{figure}

@@

 Coarsening & \# 2-cells & \texttt{phCol} & Vineyards / $d_{K}$ & Moves / $d_{\text{LCS}}$ \\

@@

Moving forward, our results suggest there are many aspects of computing persistence in dynamic settings yet to be explored. Do existing optimizations to the standard reduction algorithm have natural extensions to dynamic setting? Are there other edit distances that yield efficient algorithms to update $R = DV$ decompositions, beyond moves and transpositions? An ideal method would be one that allows arbitrary permutations to a previously computed decomposition that is provably more efficient than performing the reduction independently in all cases.

>>>>

\documentclass{siamart190516}

%\usepackage[margin=0.75in]{geometry}
\usepackage{amsmath, amssymb}
\usepackage{mathtools}
\usepackage{url}
\usepackage{enumitem}
\usepackage{setspace}
\usepackage{scalefnt}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{algorithm}
%\usepackage{algorithm2e}
%\usepackage{algorithmicx}
\usepackage[noend]{algpseudocode}
\usepackage{bbm}
\usepackage{graphicx}

@@

\newcommand\undermat[2]{%
  	\makebox[0pt][l]{$\smash{\underbrace{\phantom{%
    \begin{matrix}#2\end{matrix}}}_{\text{$#1$}}}$}#2}

@@

%\newtheorem{theorem}{Theorem}
%\newtheorem{definition}{Definition}
%\newtheorem{proposition}{Proposition}
%\newtheorem{lemma}{Lemma}
%\newtheorem{corollary}{Corollary}
% \newcommand{\CommentX}[1]{\unskip~\%~#1~\%}

@@

% Sets running headers as well as PDF title and authors
\headers{An Example Article}{D. Doe, P. T. Frank, and J. E. Smith}

% Title. If the supplement option is on, then "Supplementary Material"
% is automatically inserted before the title.
\title{Move Schedules: Fast persistence computations in sparse dynamic settings
	\thanks{Submitted to the editors DATE.
		\funding{This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758.}
	}
}

% Authors: full names plus addresses.
\author{
	Matt Piekenbrock\thanks{Department of Computational Mathematics, Science \& Engineering, Michigan State University.}
	\and Jose A. Perea\thanks{Department of Mathematics, Michigan State University.}
}

%\title{\vspace{-2.0em} Move Schedules: \\ Fast persistence computations in sparse dynamic settings\thanks{This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758.}\vspace{-0.5em}}
%\author{Matt Piekenbrock\thanks{Department of Computational Mathematics, Science \& Engineering, Michigan State University}\;\; and Jose A. Perea$^\dagger$\thanks{Department of Mathematics, Michigan State University}}
%\date{}

% Optional PDF information
\ifpdf
\hypersetup{
  pdftitle={An Example Article},
  pdfauthor={Matt Piekenbrock, P. T. Frank, and J. E. Smith}
}
\fi

% ------  Begin document ------- 
\begin{document}

\maketitle

% REQUIRED - "Abstracts must be able to stand alone and so cannot contain citations to the paper's references, equations, etc. 
\begin{abstract}
	The standard procedure for computing the persistent homology of a filtered simplicial complex is the matrix reduction algorithm. Its output is a particular decomposition of the total boundary matrix, from which the persistence diagrams and generating cycles can be derived. 
	Persistence diagrams are known to vary continuously with respect to their input; this motivates the algorithmic study of persistence computations for time-varying filtered complexes. Computationally, simulating persistence dynamically can be reduced to maintaining a valid decomposition under adjacent transpositions in the filtration order. In practice, the quadratic scaling in the number of  transpositions often makes this maintenance procedure slower than simply computing the decomposition from scratch, effectively limiting the application of dynamic persistence to relatively small data sets. In this work, we propose a coarser strategy for maintaining the decomposition over a discrete 1-parameter family of filtrations. Our first result is an analysis of a simple linear-time strategy which reduces the number of column operations needed to simulate persistence across a fixed homotopy by at most a factor of 2. We then show a modification of this technique which maintains only a sublinear number of valid states, as opposed to a quadratic number of states, and we provide tight lower bounds for this technique.
	Finally, we provide empirical results suggesting that the decrease in operations needed to compute diagrams across a family of filtrations is proportional to the difference between the expected quadratic number of states, and the proposed sublinear coarsening.
\end{abstract}

% REQUIRED
\begin{keywords}
  example, \LaTeX
\end{keywords}

% REQUIRED
\begin{AMS}
  68Q25, 68R10, 68U05
\end{AMS}

@@

 \subsection{A Motivating Example and Contributions}\label{sec:motivation} 
As a motivating example, we describe a simple experiment illustrating why the vineyards algorithm does not always yield an efficient method in many time-varying applications. 
Consider a series of grayscale images (i.e. a video) depicting a fixed-width annulus expanding about the center of a $9 \times 9$ grid, where scale of the annulus is parameterized by a single parameter $\tau$. Thus, the data set is a 1-parameter family of images. Each image contains the same number of pixels whose intensities---these are determined by the parameter---vary with time. For each image, we build a filtered simplicial complex using the \emph{Freudenthal} triangulation, with the ordering of inclusions obtained by lower stars.  
 \begin{figure*}[t]
 	\centering
 	\includegraphics[width=0.36\textwidth, height = 1.45in]{circle_vineyards.png}
 	\includegraphics[width=0.63\textwidth, height = 1.45in]{circle_complex.png}
 	\caption{On the right, the grayscale image data set and corresponding lower star filtrations. On the left, the cumulative number of column operations required to compute persistence for this time-varying filtration. The persistent Betti numbers for each filtration from left to right are: $\beta_1 = (1,0,0), \beta_2 = (1,1,0), \beta_3 = (1,1,0), \beta_4 = (1,1,0), \beta_5 = (4,0,0)$. Observe computing $10$ diagrams independently at evenly spaced time points (the green line) both captures these major topological changes and is the most computationally efficient approach shown.}
 	\label{fig:vineyards}
 \end{figure*}
Snapshots of the image data and its corresponding time-varying filtration is depicted in Figure~\ref{fig:vineyards} (top right and bottom right, respectively), which we use as a baseline for comparing the standard algorithm $\texttt{pHcol}$ with vineyards. Note the size of the filtration is fixed--no simplices are ever added or removed over time. 
There are two events which dramatically alter the persistence diagrams extracted from these filtrations: the first occurs when the central connected component splits to form a cycle, and the second occurs when the annulus approaches the outer dimensions of the grid, splitting into four components. As a consequence, in this example, only a few persistence diagrams are needed to capture the major topological changes over time.

@@

\textbf{Our results and approach are as follows:} First, we leverage work of Busaryev et. al.~\cite{busaryev2010tracking} to include a coarser type of operation to the dynamic setting, called \emph{moves}. We give a tight lower bound on the number of moves needed to perform an arbitrary permutation to the $R = D V$ decomposition, and give a proof of optimality by a reduction to the permutation edit distance problem.
We also give worst-case bounds in expectation as well as efficient algorithms for achieving these bounds---both of which are derived from a reduction to the Longest Increasing Subsequence (LIS) problem. 
This reduction yields an efficient algorithm for generating sequences of moves $( s_1, s_2, \dots, s_d )$, which we call \emph{schedules}. 
We investigate the feasibility of optimizing the cost 
(i.e., number of column operations) of these schedules directly, including the existence of a greedy-type approach, which we show can lead to arbitrarily bad behavior. 
In light of these results, we give an alternative proxy-objective to minimize, provide bounds justifying its relevance to the original objective, and give an efficient $O(d \log d)$ algorithm for approximately solving this proxy minimization. 
A performance comparison with other reduction-based persistent homology computations is given, wherein our approach is demonstrated to be an order of magnitude more efficient than existing approaches in applications such as flock simulation and 2D persistence computation. We conclude with a discussion of situations where our strategy is most applicable and future directions for research.

@@

 The main output of the reduction algorithm is a matrix decomposition $R = D V$, where the persistence diagram is encoded in $R$. 
 Given a  filtration $K$ with $m$ simplices and maximal dimension $d$, 
 one assembles the elementary boundary chains $\partial(\sigma)$ as columns ordered according to the filtration order, forming the \emph{filtration boundary matrix} $D$. 
 The reduction algorithm can  compute the persistent homology for all dimensions up to $d - 1$.
 In this case,  $D$ is a square matrix of size $m \times m$. Alternatively, a single dimension $i \leq d - 1$ can be computed, in which case one reduces a pair of matrices $D_i, D_{i+1}$, where the former has dimension $d_{i-1} \times d_i$ and the latter $d_{i} \times d_{i+1}$. If one is interested in a particular dimension of homology then the latter approach is preferred. However, the algorithmic description of the algorithms is simpler when considering the full matrix, so $D$ will always be $(m \times m)$ in the descriptions that follow. All algorithms discussed generalize to both situations. We give an example using the matrix pair $(R_1, V_1)$  below. 
\\
\\
\noindent
\textbf{Example 2.1: Reduction} Consider a triangle with vertices $u,v,w$ and edges $a = (u,w)$, $b = (v,w)$, $c = (u, v)$ whose filtration order, read as inclusions from left to right, is given as $K = (u, v, w, a, b, c)$. The reduction algorithm begins by setting $R = D$ and $V = I$, followed by left-to-right column operations until invariant (2) below is satisfied.
If the $i$-th column of $R$ is nonzero, then $\mathrm{low}_R(i)$ 
is the row index of its lowest nonzero entry.
Using $\mathbb{Z}_2$ coefficients and homology in dimension 1 to simplify the presentation, the steps look at follows:
\begin{displaymath}
	\def\rowsp{0.4em}
	\def\arraystretch{0.95}
	\makeatletter\setlength\BA@colsep{1.2pt}\makeatother
	%
	\begin{blockarray}{cccc}{R_1}
	 \kern-0.2em & a & b & c  \\
		\begin{block}{c[ccc]}
  		u \kern\rowsp  & \; 1 &    &  1 \; \topstrut \\
  		v \kern\rowsp & \;     &  1 & 1 \; \\
  		w \kern\rowsp & \; 1 &  1 &    \; \botstrut \\
		\end{block}
	\end{blockarray}
	\; , \;
	\begin{blockarray}{cccc}{V_1}
	\kern-0.2em & a & b & c  \\
		\begin{block}{c[ccc]}
        a \kern\rowsp  & \; 1 &  &  \; \topstrut \\
  		b \kern\rowsp & \;  & 1 & \; \\
  		c \kern\rowsp & \;  &  & 1 \; \botstrut \\
		\end{block}
	\end{blockarray}
	\quad \xrightarrow{} \,
	\begin{blockarray}{cccc}{R_1}
	\kern-0.2em & a & b & c  \\
		\begin{block}{c[ccc]}
  		u \kern\rowsp  & \; 1 & 1 & 1 \; \topstrut \\
  		v \kern\rowsp & \;     & 1 & 1 \; \\
  		w \kern\rowsp & \; 1 &    &    \; \botstrut \\
		\end{block}
	\end{blockarray}
	\; , \;
	\begin{blockarray}{cccc}{V_1}
	\kern-0.2em & a & b & c  \\
		\begin{block}{c[ccc]}
  		a \kern\rowsp  & \; 1 & 1 &    \; \topstrut \\
  		b \kern\rowsp & \;     & 1 &    \; \\
  		c \kern\rowsp & \;     &    & 1 \; \botstrut \\
		\end{block}
	\end{blockarray}
	\quad \xrightarrow{} \,
	\begin{blockarray}{cccc}{R_1}
	\kern-0.2em & b & a & c  \\
		\begin{block}{c[ccc]}
  		u \kern\rowsp  & \;     & 1  &  \; \topstrut \\
  		v \kern\rowsp &  \; 1  & 1  &  \; \\
  		w \kern\rowsp & \; 1  &     &  \; \botstrut \\
		\end{block}
	\end{blockarray}
	\; , \;
	\begin{blockarray}{cccc}{V_1}
	\kern-0.2em & b & a & c  \\
		\begin{block}{c[ccc]}
  		b \kern\rowsp  & \; 1 & 1 & 1 \; \topstrut \\
  		a \kern\rowsp & \;     & 1 & 1 \; \\
  		c \kern\rowsp & \;     &    & 1 \; \botstrut \\
		\end{block}
	\end{blockarray}
	\vspace*{-1em}
\end{displaymath}

@@

%\begin{proposition}[Persistent Pairing~\cite{zomorodian2005computing}]
%	Let $R = D V$, where $R$ is in reduced form and $V$ is full-rank upper triangular (invariants (1) and (2)  hold). Then the pairing $(\sigma_i, \sigma_j)$ defined by $i = \mathrm{low}_R(j)$ is a \emph{creator/destroyer (persistent) pair}. While the matrices $R,V$ satisfying (1) and (2) are not unique in general,  the persistent pairs $(i,j)$ are unique and define the persistence diagrams of $K$.
%\end{proposition}

@@

	\def\rowsp{0.4em}

@@

	\makeatletter\setlength\BA@colsep{1.2pt}\makeatother

@@

  		u \kern\rowsp  & \; 1 & 1 &  \; \topstrut \\
  		v \kern\rowsp & \;  & 1 &  \; \\
  		w \kern\rowsp & \; 1 &  &  \; \botstrut \\

@@

	\,\,, 

@@

        a \kern\rowsp  & \; 1 & 1 & 1 \; \topstrut \\
  		b \kern\rowsp & \;  & 1 & 1 \; \\
  		c \kern\rowsp & \;  &  & 1 \; \botstrut \\

@@

	\hspace{-0.1em}
	\begin{blockarray}{cccc}{}

@@

  		 & \; 1 &  &  \; \topstrut \\
  		 & \;  & 1 &  \; \\
  		 & \; 1 & 1 & \; \botstrut \\

@@

	\,\,, 
	\begin{blockarray}{cccc}{}

@@

  		 & \; 1 &  & 1 \; \topstrut \\
  		 & \;  & 1 & 1 \; \\
  		 & \;  &  & 1 \; \botstrut \\

@@

	\hspace{-0.2em}
	\begin{blockarray}{cccc}{}
  		& \;  & 1 &  \; \topstrut \\
  		& \; 1 &  &  \; \\
  		& \; 1 & 1 &  \; \botstrut \\
	\,\,, 
	\begin{blockarray}{cccc}{}
  		& \; 1 &  & 1 \; \topstrut \\
  		& \;  & 1 & 1 \; \\
  		& \;  &  & 1 \; \botstrut \\
	\hspace{-0.1em}

@@

  		u \kern\rowsp  & \;  & 1 &  \; \topstrut \\
  		v \kern\rowsp & \; 1 & 1 &  \; \\
  		w \kern\rowsp & \; 1 &  &  \; \botstrut \\

@@

  		b \kern\rowsp  & \; 1 & 1 & 1 \; \topstrut \\
  		a \kern\rowsp & \;  & 1 & 1 \; \\
  		c \kern\rowsp & \;  &  & 1 \; \botstrut \\

@@

If one has a pair of filtrations $(K_0, K_1)$ each with $m$ simplices, and we assume that each simplex $\sigma_i$ switches its relative ordering with another simplex $\sigma_j$ at most once in some ordered schedule of transpositions, then the number of intermediate persistence diagrams is bounded above by $O(m^2)$ using transpositions, whereas with move operations this bound clearly is at most $O(m)$.  
\\
\\
\noindent \textbf{Example:} We continue the example used in sections~\ref{sec:vineyards} and~\ref{sec:reduction} to illustrate moves. Consider moving the edge $a$ to the position taken by the edge $c$ in the filtration. The donor columns introduced in~\cite{busaryev2010tracking} are shown on the left side of each matrix. Note that using vineyards, the equivalent permutation to the decomposition would require $4$ column operations on both $R_1$ and $V_1$, respectively, whereas a single move operation accomplishes this permutation using only $2$ column operations per matrix. 
\begin{displaymath}
	\def\rowsp{0.4em}
	\def\arraystretch{0.95}
	\makeatletter\setlength\BA@colsep{1.2pt}\makeatother
	%
	\begin{blockarray}{cc}{d_R}
	\kern-0.2em & a \\
		\begin{block}{c[c]}
  		u \kern \rowsp & \; 1 \; \topstrut \\
  		v \kern \rowsp & \;  \; \\
  		w \kern \rowsp & \; 1 \; \botstrut \\
		\end{block}
	\end{blockarray}
	\; \;
	\begin{blockarray}{cccc}{R_1}
	 \kern-0.2em & a & b & c  \\
		\begin{block}{c[ccc]}
  		u \kern \rowsp & \; 1 & 1 &  \; \topstrut \\
  		v \kern \rowsp & \;  & 1 &  \; \\
  		w \kern \rowsp & \; 1 &  &  \; \botstrut \\
		\end{block}
	\end{blockarray}
	%
\; \to
	%
\begin{blockarray}{cc}{}
	\kern-0.2em & b \\
		\begin{block}{c[c]}
  		  & \; 1 \; \topstrut \\
  		 & \;  1 \; \\
  		 & \; \; \botstrut \\
		\end{block}
	\end{blockarray}
	\; \;
\begin{blockarray}{cccc}{}
	\kern-0.2em & a & b & c  \\
		\begin{block}{c[ccc]}
  		  & \; 1 &    &  \; \topstrut \\
  		  & \;    & 1 &  \; \\
  		 & \; 1 & 1 &  \; \botstrut \\
		\end{block}
	\end{blockarray}
	%
\; \to
	%
	\begin{blockarray}{cc}{}
	\kern-0.2em & c \\
		\begin{block}{c[c]}
  		  & \; \hphantom{1} \; \topstrut \\
  		 & \;  \hphantom{1} \; \\
  		 & \; \hphantom{1} \; \botstrut \\
		\end{block}
	\end{blockarray}
	\; \; 
	\begin{blockarray}{cccc}{}
	\kern-0.2em & a & b &  c \\
		\begin{block}{c[ccc]}
  		  & \; 1   &    &   1  \; \topstrut \\
  		 & \;       &  1  &  1  \; \\
  		 & \; 1   &   1   &     \; \botstrut \\
		\end{block}
	\end{blockarray}
%
\; \xrightarrow{P} 
\begin{blockarray}{cc}{}
	\kern-0.2em & c \\
		\begin{block}{c[c]}
  		  & \; \hphantom{1} \; \topstrut \\
  		 & \; \hphantom{1} \; \\
  		 & \; \hphantom{1} \; \botstrut \\
		\end{block}
	\end{blockarray}
\; \;
\begin{blockarray}{cccc}{}
	\kern-0.2em & b & c & a  \\
		\begin{block}{c[ccc]}
  		  & \;  & 1 &  1\; \topstrut \\
  		 & \; 1 & 1 &  \; \\
  		 & \; 1 &  & 1 \; \botstrut \\
		\end{block}
	\end{blockarray}
%
\; \xrightarrow{d_R} 
\begin{blockarray}{cccc}{R_1}
\kern-0.2em & b & c & a  \\
	\begin{block}{c[ccc]}
		u \kern \rowsp & \;  & 1 &  \; \topstrut \\
		v \kern \rowsp & \; 1 & 1 &  \; \\
		w \kern \rowsp & \; 1 &  &  \; \botstrut \\
	\end{block}
	% \undermat{B}{ & & }  
\end{blockarray}
\end{displaymath}

@@

\begin{figure}[t]
	\centering
	\includegraphics[width=0.95\textwidth]{crocker_combo_1.png}
	\caption{An example of a crocker plot (right) depicting the evolution of dimension $p = 1$ Betti curves over time. The green $\mathrm{X}$ marks correspond chronologically to the complexes (left), in row-major order. The large orange and purple areas depict $1$-cycles persisting in both space (the y-axis) and time (x-axis). }\label{fig:crocker1}
\end{figure}

@@

\small{Coarsening} & \# 2-cells & \texttt{phCol} & Vineyards / $d_{K}$ & Moves / $d_{\text{LCS}}$ \\

@@

Moving forward, our results suggest there are many aspects of computing persistence in dynamic settings yet to be explored. For example, it's not immediately clear whether one could adopt, for example, the twist optimization used in the reduction algorithm to the dynamic setting. Another avenue of interest is the setting where one has knowledge a priori of the time-varying... particular spaces
